An adaptive parallel algorithm for Vector Error Correction models applied to the Foreign Exchange Market

### 2016-07-06 articles-conferences/iscef2016/old-manuscript/manuscript.tex
Final manuscript

\chapter{Adaptive parallel algorithm for Vector error correction model}

\vspace{0.5cm}

Vector error correction model (VECM) parameters are obtained using ordinary
least squares (OLS) method \cite{golub1980}, which is computationally expensive
when the number of past values $L$ and observations $p$ increases. Moreover,
obtaining cointegration vectors is also an expensive routine because of the use
of the Johansen method.

In this study we found that for some combinations of $L$ and $p$, the number of
cointegration relations is increased and this leads to better performance. We
propose to parallelise this search in order to achive faster execution times.

\section{The problem} \label{sec:proposal}

Cointegration vectors can be found applying the Johansen method which uses a
sample of the most recent historical data. However, one of the main limitations of this
methodology is that it assumes cointegration vectors do not change in time.  In
fact, the long-run relationships between the time series might change due to
several economic factors that can lead to structural breaks in the cointegrating
relation \cite{gregoryETal1996}.

In this study, we show that cointegration relations depend on the amount $L$ of
historical data considered, and how many lags $p$ were used. This search is
computationally expensive, considering that the Johansen
method~\cite{johansen1995} (which is of order $O(n^3)$) is used and these
parameters change at every iteration.

Our proposal consists on to parallelise this search in order to get new
parameters at every iteration.

In finance, little academic reasearch has been done using high-frequency Forex
rates \cite{Bekiros2015}. In our study, tests were run using four currency
rates: Euro (EUR) to United States Dollar (USD) (EURUSD), British Pound (GBP) to
USD (GBPUSD), USD to Swiss Franc (CHF) (USDCHF) and USD to Japanese Yen (JPY)
(USDJPY) with a 10-seconds frequency.  The forecast capability of our method was
later measured using MAPE, MAE and RMSE \cite{armstrong1992}.

\section{Methodology}

VECM is obtained re-writing the VAR model, shown in equation (\ref{eq:var}), in
terms of differences defines as: $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
The VECM model, expressed in terms those differences, takes the form:
\begin{equation}\label{eq:vec}
\Delta \mathbf{y}_t 
= \boldsymbol{\Omega}\,\mathbf{y}_{t-1}
  + \sum_{i=1}^{p-1} \boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
where the coefficients matrices $\boldsymbol{\Phi}_i^*$ and 
$\boldsymbol{\Omega}$, expressed in terms of the matrices
$\boldsymbol{\Phi}_i$ of (\ref{eq:var}), are:
\begin{align*}
\boldsymbol{\Phi}_i^* 
&:= -\sum_{j=i+1}^{p}\boldsymbol{\Phi}_j\,, \\
\boldsymbol{\Omega}
&:= -\left( \mathbb{I} - \boldsymbol{\Phi}_1 - \dots 
    - \boldsymbol{\phi}_p \right)\,. 
\end{align*}
The following well known properties of the matrix $\boldsymbol{\Omega}$
\cite{johansen1995} will be useful in the sequel:
\begin{itemize}
\item
If $\boldsymbol{\Omega} = \mathbf{0}$, there is no cointegration.
\item 
If $rank(\boldsymbol{\Omega})=l$, i.e., if $\boldsymbol{\Omega}$ has
full rank, then the time series are not I(1) but stationary.
\item
If $rank(\boldsymbol{\Omega})=r$, $0<r<l$, then there is cointegration
and the matrix $\boldsymbol{\Omega}$ can be expressed as
$\boldsymbol{\Omega}=\boldsymbol{\alpha\beta}^\top$, where $\boldsymbol{\alpha}$
and $\boldsymbol{\beta}$ are
$l\times r$ matrices and
$\text{rank}(\boldsymbol{\alpha})=\text{rank}(\boldsymbol{\beta})=r$.
\item
The columns of $\boldsymbol{\beta}$ contains the cointegration vectors and the rows of
$\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
$\boldsymbol{\beta}$ is obtained by Johansen procedure~\cite{johansen1988},
whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.
\end{itemize}

If cointegration exists, then equation (\ref{eq:vec}) can be written
as follows:
\begin{equation}\label{eq:vecfull}
\Delta\mathbf{y}_t 
= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
which is a VAR model but for time series differences.

@@

\subsection{Motivation}
The relation between the number of cointegration relations
and parameters $L$ and $p$ is studied and then a relation between cointegration
and performance is stated.

Figure~\ref{fig:hists} shows the number $r$ of cointegration vectors given by
the Johansen method for different values of the parameters $L$ leaving $p$ set
as $1$, using four currencies data, i.e., $l=4$, and 1000 considering iterations
($it=1000$).
\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=\textwidth]{img/histCointVectors}
  \caption{Histogram of the number of cointegration vectors using $p=1$. Four
  possible values for $L$ are shown (100, 1900, 6100 and 10700).
  In the cases of $L=100$ and $L=6100$, most of the time there is little
  cointegration (0 or 4 cointegration vectors). However, for $L=1900$ and
  $L=10700$ significant cointegration is found.}
  \label{fig:hists}
\end{figure}
We know that when no cointegration vectors can be found, i.e $r=0$, implies that
the variables are not cointegrated. In the same way, when $r=l$ reveals that
none of the processes is I(1) but, they are all stationary.  The interesting cases of
cointegration are those, where $r$ lies strictly between $0$ and $l$, i.e.
$0<r<l$,

@@

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=\textwidth]{img/histCointVectorsp5}
  \caption{Histogram of the number of cointegration vectors using $p=5$. 
   Despite the change in the distribution of cointegration vectors, 
   the percentage of cointegration is nearly the same as in the case $p=1$. 
   For example, for $L=10700$ the sum of 1 and 2 number of cointegration
   vectors is very similar to 1 cointegration vector for the case $L=10700$
   in figure \ref{fig:hists}.}
   \label{fig:histsp5}
\end{figure}

The same experiments, setting $p=5$, were carried out in order to see the
effect on the number of cointegration vectors found. 
Figure~\ref{fig:histsp5} shows that, despite the change in the distribution
of the number of cointegration vectors, the sum of the cases where $0<r<4$,
remains nearly the same. 
Therefore, we can say that the number of lags doesn't significantly affect
the extent of cointegration found.

@@

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.7\textwidth]{img/MAPEvsCoint-offset21600-p-2-freq-10s}
  \caption{MAPE versus percentage of cointegration. MAPE was obtained from 1000
  iterations considering several possible $L=[100,10000,400]$ and $p=[1,5]$.
  Best $L$ found was 8900.}
  \label{fig:cointvsmape}
\end{figure}

From these results we observe that cointegration changes with different values
for $L$ and $p$ and also that better cointegration percentage leads to better
accuracy performance.  Therefore we propose to choose best $L$ and $p$
parameters determined at every step.  The criterion for choosing $L$ and $p$ is
to maximise the percentage of cointegration $PC$ considering a number of past
iterations $it$.

Algorithm ~\ref{alg:AVECM} solves equation \ref{eq:vareq} defining first
parameters $L$ and $p$ using the function {\bf get\_best\_params} which makes a
grid search on the two vector lists $Ls$ and $ps$ and returns the parameters $L$
and $p$ which maximise the percentage of cointegration $PC$ (see
equation~\ref{eq:pcoint}) for a pre-defined number of iterations.  Since this
search is computational expensive, the number of cointegration vectors at each
iteration is computed in parallel, thus ensuring a response before new data is
available.  This parallel routine also allows to refine our grid search leading
to obtain improved forecasting performance.  After that optimum $L$ and $p$
parameters are found, VECM is built and used to forecast next data.

@@

@@

\begin{table}[h!]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
& \textbf{Statistic} & \textbf{Critical value} & \textbf{Result}\\
\hline
EURUSD          &  -0.05398   & -1.94101 & True       \\
$\Delta$ EURUSD & -89.12344   & -1.94101 & False       \\
GBPUSD          &  -0.88208   & -1.94101 & True          \\
$\Delta$ GBPUSD & -82.56501   & -1.94101 & False       \\
CHFUSD          & -0.49941    & -1.94101 & True         \\
$\Delta$ CHFUSD & -74.21940   & -1.94101 & False       \\
JPYUSD          &  0.36721    & -1.94101 & True        \\
$\Delta$ JPYUSD & -101.15589  & -1.94101 & False     \\ 
\hline
\end{tabular}
\end{center}
\caption{Unit roots tests for EURUSD, GBPUSD, USDCHF and USDJPY at 10-seconds
frequency.}
\label{tab:adf}
\end{table}

\section{Parallel implementation} \label{sec:paralell}

@@

\begin{figure}[!h]
  \centering
  \includegraphics[width=\textwidth]{img/accuracy}
  \caption{AVECM out-of-sample forecasting for four currency rates.}
  \label{fig:accuracy}
\end{figure}

\begin{table*}[ht!]
\tiny
\caption{Performance and execution times for min MAPE criteria}
\label{tab:stats}
\begin{center}
\begin{tabular}{|l|l|l|c|c|c|c|c|}
\hline
\multicolumn{3}{|c}{Parameters} & 
\multicolumn{3}{|c}{Out-of-sample Performance} &
\multicolumn{2}{|c|}{Time[s]} \\ 
\hline

@@

\begin{figure}[!h]
  \centering
  \includegraphics[width=0.7\textwidth]{img/extimes}
  \caption{Computing time of sequential and parallel algorithm is shown in the
  upper figure. Speed-up is shown below.}
  \label{fig:extimes}
\end{figure}

Execution times do not consider the loading data time, just the finding of
best parameters and the building the VECM model. The time MPI spends
transferring data and synchronising process is about two seconds independently
of the number of process considered.

>>>>

\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
\usepackage{amsfonts,amssymb,amsmath,
            amssymb,amsopn,amsxtra,amstext,amsbsy}
\usepackage{epsfig}
\usepackage[makeroom]{cancel}
\usepackage{array,arydshln}
\usepackage{array}
\usepackage{url}
%\usepackage{tikz}
\usepackage[utf8x]{inputenc}
\usepackage{fancyhdr}
\usepackage{hyperref,wasysym}
\usepackage{color}
\usepackage{enumerate}
\usepackage{algorithm}
\usepackage{algorithmic}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\usepackage{multirow}
%\usepackage[spanish]{babel}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{xcolor}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{hyperref}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{pmat}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\newtheorem*{myteo}{Teorema}
%\newtheorem*{mydef}{Definici\'on}
%% MYREMARK ENVIRONMENT %%%%%%%%%%%
\newenvironment{mycolor}{\color{blue}}{}
\newcounter{myremarkcount}
\setcounter{myremarkcount}{0}
\newsavebox{\myremarkbox}
\newenvironment{myremark}
{% This is the begin code
\medskip
\begin{flushright}
\begin{lrbox}{\myremarkbox}%
\begin{minipage}[t]{1.00\linewidth}
\stepcounter{myremarkcount}
\footnotesize
\begin{mycolor}
{{\bf Proofreading Remark.}\ }
% \begin{it}
}
{% This is the end code
% \end{it}
\end{mycolor}
\end{minipage}
\end{lrbox}
\fbox{\usebox{\myremarkbox}}
\end{flushright}

\bigskip}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\journalname{Cluster Computing}

\begin{document}

\title{An adaptive parallel algorithm for Vector Error Correction models applied
to the Foreign Exchange
Market
}
%\subtitle{Do you have a subtitle?\\ If so, write it here}

%\titlerunning{Short form of title}        % if too long for running head

\author{Paola Arce \and Jonathan Antognini \and Werner Kristjanpoller \and Luis Salinas }

%\authorrunning{Short form of author list} % if too long for running head

\institute{P. Arce  \at
              Departamento de Inform\'atica, Universidad T\'ecnica Federico Santa
              Mar\'ia\\
              Tel.: +56-32-2654000\\
              %Fax: +123-45-678910\\
              \email{paola.arce@usm.cl}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
           J. Antognini \at
              Departamento de Inform\'atica, Universidad T\'ecnica Federico Santa
              Mar\'ia\\
              %Fax: +123-45-678910\\
              \email{jantognini@csrg.cl}             \\
           \and
           W. Kristjanpoller \at
              Departamento de Industrias, Universidad T\'ecnica Federico Santa
              Mar\'ia\\
              %Fax: +123-45-678910\\
              \email{werner.kristjanpoller@usm.cl}             \\
           \and
           L. Salinas \at
              Departamento de Inform\'atica, Universidad T\'ecnica Federico Santa
              Mar\'ia\\
              Centro Científico Tecnológico de Valparaíso (CCTVal) \\
              \email{luis.salinas@usm.cl}           %  \\
}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}

A Vector Error Correction Model (VECM) is an econometric model which
characterizes the joint dynamic behaviour of a set of cointegrated variables. It
is said that two non-stationary time series are cointegrated if a linear
combination of them exhibit a stationary behaviour.

Cointegration is commonly found in finance and it has been shown that
introducing cointegration restrictions often improves forecasting. However,
determining VECM parameters and how much of the historical data is useful for
making a forecast is a computational challenge.

In this study, we propose a new version of VECM whose parameters are obtained
adaptively using parallel computing taking into account the cointegrated
behaviour of the time series. The methodology is tested using data of the
Foreign Exchange (Forex) market. Numerical experiments show that our adaptive
and parallel approach significantly improves execution times with respect to the
common serial processing and forecasts are made before new data arrives.

\keywords{VECM \and Cointegration \and Forex \and MPI \and Parallel algorithm}

\end{abstract}

\section{Introduction}
\label{sec:introduction}
In finance, it is common to find variables with long-run equilibrium
relationships. This is called cointegration and it reflects the idea of that
some sets of variables cannot wander too far from each other. Cointegration
means that one or more linear combinations of these variables are stationary
even though individually they are not. Furthermore, the number of cointegration
vectors reflects how many of these linear combinations exist.  Some models,
such as the Vector Error Correction (VECM) \cite{engle87}, take advantage of
this property and describe the joint behaviour of several cointegrated
variables.

VECM introduces this long-run relationship among a set of cointegrated
variables as an error correction term. VECM is a special case of the vector
autoregressive model (VAR) model \cite{sims1980}. VAR model expresses future
values as a linear combination of variables past values.  However, the VAR
model cannot be used with non-stationary variables. VECM is a linear model but
in terms of variable differences. If cointegration exists, variable differences
are stationary and they introduce an error correction term which adjusts
coefficients to bring the variables back to equilibrium.

Both VECM and VAR model parameters are obtained using ordinary least squares
(OLS) method~\cite{golub1980}. Since OLS involves many calculations, the parameter estimation
method is computationally expensive when the number of past values and
observations increases. Moreover, obtaining cointegration vectors 
is also an expensive routine.

In finance, many economic time series are revealed to be stationary when they
are differentiated and moreover cointegration restrictions often improve
forecasting \cite{duy1998}. Therefore, VECM has been widely adopted in
finance: \cite{mukherjee1995}, \cite{maysami2000}, \cite{arestis2001} and
\cite{seong2013} to name a few. Pair
trading is a very common example of a cointegration
application \cite{herlemont2003} but also can also be extended to a larger set
of variables \cite{mukherjee1995}, \cite{engle2004}.

Based on this fact, we proposed in this paper a new forecast method which
determines adaptively the number of observations and lags of VECM based on
cointegration which we called Adaptively Vector Error Correction (AVECM).  Since
cointegration often improves forecasting, model parameters will be those which
maximize cointegration relations in the past.  This search is computationally
expensive because to find cointegration vectors we used the Johansen
method~\cite{johansen1995}, which is of order $O(n^3)$. Therefore, our proposal
is to parallelize this search in order to get new parameters at every iteration.

Model effectiveness is focused on out-of-sample forecast rather than in-sample
fitting. This criterion allows our proposal AVECM prediction capability to be
expressed rather than just explaining data history.

The forecast capability of our method was measured using MAPE, MAE and
RMSE~\cite{armstrong1992}. Tests were run using four currency rates: Euro (EUR) to
United States Dollar (USD) (EURUSD), British Pound (GBP) to USD (GBPUSD), USD to Swiss
Franc (CHF) (USDCHF) and USD to Japanese Yen (JPY) (USDJPY) with a 10-seconds frequency.

This paper is organized as follows: section~\ref{sec:background} presents the
VAR and VECM, the AVECM algorithm is presented in section~\ref{sec:methodology}.
In section~\ref{sec:results} we describe the tests carried on to assess the
accuracy and the execution time of AVECM.  This section also includes a
description of the test data.  Section~\ref{sec:conclusions} contains the
conclusions and a discussion of future research.

\section{Background}
\label{sec:background}
\subsection{Integration and Cointegration}\label{sec:coint}\  
Following Johansen \cite{johansen1995} we shall say that a stochastic process
$Y_t$ which satisfies $Y_t-E(Y_t) = \sum_{i=0}^\infty C_i\,\varepsilon_{t-i}$ is
called $I(0)$, and then we shall write $Y_t\sim I(0)$, whenever
$\sum_{i=0}^\infty C_i \neq 0$ and $\sum_{i=0}^\infty C_i\,z^i$ converges for
$z\in\mathbb{C}$ with $|z|<1$.  It is understood that the condition
$\varepsilon_t\sim iid(0,\sigma^2)$ holds.

A (vector) time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
$d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
(discrete) differentiation an stationary process is obtained~\cite{banerjee1993};
more precisely, whenever
$(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
$(1-L)\,\mathbf{y}_t = \Delta\mathbf{y}_t = \mathbf{y}_t-\mathbf{y}_{t-1}$ for
all $t$.

Note that this definition includes the scalar case as time series of
vectors of dimension 1; in this scalar case we will write the time series in
nonbold format.

Let $\mathbf{y}_t^\nu$, $\nu=1,\dots,p$, be a set of $p$ vector time series of
order $I(1)$.  They are said to be {\em cointegrated\/} if a vector
$\beta=[\beta(1),\dots,\beta(p)]^\top \in \mathbb{R}^p$ exists, such that the
time series,
\begin{equation}
\mathbf{Z}_t:= 
\sum_{\nu=1}^p \beta(\nu)\,\mathbf{y}_t^\nu\,\sim\,\text{I(0)}\,.
\end{equation}
In other words, a set of $I(1)$ time series is said to be cointegrated if a
linear combination of them exists, which is I(0).

\subsection{Vector Autorregresive Models}\label{sec:varvec}

VECM is a special case of VAR model and both describe the joint behaviour
of a set of variables.

The VAR($p$) model is a general framework describing the behaviour of a
set of $l$ endogenous variables as a linear combination of their last $p$
values, where $l,p\in\mathbb{N}$. 
In our case, each one of these $l$ variables is a scalar time series
$y_{\lambda,t}$, $\lambda=1,\dots,l$, and we represent them all together
at time $t$ by the the vector time series:
\begin{equation}
\label{eq:variables}
\mathbf{y}_t = 
\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t} \end{bmatrix}^\top.
\end{equation}
\noindent
Notice that the vectors $\mathbf{y}_t$ are assumed to be $l$-dimensional.

The VAR($p$) model describes the behaviour of a dependent variable in terms of
its own lagged values and the lags of the others variables in the system. The
model with $p$ lags is formulated as the system of $N$:
\begin{align}
\label{eq:var}
\mathbf{y}_t 
= \boldsymbol{\Phi}_1 \mathbf{y}_{t-1} +
  \boldsymbol{\Phi}_2 \mathbf{y}_{t-2} + \dots +
  \boldsymbol{\Phi}_p\mathbf{y}_{t-p} +
  \mathbf{c} + \boldsymbol{\epsilon}_t \nonumber \\
t=p+1,\dots,N,
\end{align}
\noindent where 
$\boldsymbol{\Phi}_1, \boldsymbol{\Phi}_2,\dots,\boldsymbol{\Phi}_p$
are $l\times l$-matrices of real coefficients,
$\boldsymbol{\epsilon}_{p+1},
 \boldsymbol{\epsilon}_{p+2}, \dots, \boldsymbol{\epsilon}_N$ 
are error terms, $\mathbf{c}$ is a constant vector and $N$ is the total
number of samples.

Notice that, regarding our notation of section (\ref{sec:coint}),
we have here 
$\mathbf{y}_t^0 = \mathbf{y}_t$,
$\mathbf{y}_t^\nu = \mathbf{y}_{t-\nu}$ and
the $\lambda$-th component of the vector time series $\mathbf{y}_t^\nu$
is the scalar time series $y_{\lambda,t}^\nu$, where $\nu=1,\dots,p$ and
$\lambda=1,\dots,l$.

VECM is a special form of a VAR model for I(1) variables that are also
cointegrated~\cite{banerjee1993}.

It is obtained re-writing equation (\ref{eq:var}) in terms of the new
variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
The VECM model, expressed in terms those differences, takes the form:
\begin{equation}\label{eq:vec}
\Delta \mathbf{y}_t 
= \boldsymbol{\Omega}\,\mathbf{y}_{t-1}
  + \sum_{i=1}^{p-1} \boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
where the coefficients matrices $\boldsymbol{\Phi}_i^*$ and 
$\boldsymbol{\Omega}$, expressed in terms of the matrices
$\boldsymbol{\Phi}_i$ of (\ref{eq:var}), are:
\begin{align*}
\boldsymbol{\Phi}_i^* 
&:= -\sum_{j=i+1}^{p}\boldsymbol{\Phi}_j\,, \\
\boldsymbol{\Omega}
&:= -\left( \mathbb{I} - \boldsymbol{\Phi}_1 - \dots 
    - \boldsymbol{\phi}_p \right)\,. 
\end{align*}
The following well known properties of the matrix $\boldsymbol{\Omega}$
\cite{johansen1995} will be useful in the sequel:
\begin{itemize}
\item
If $\boldsymbol{\Omega} = \mathbf{0}$, there is no cointegration.
\item 
If $rank(\boldsymbol{\Omega})=l$, i.e., if $\boldsymbol{\Omega}$ has
full rank, then the time series are not I(1) but stationary.
\item
If $rank(\boldsymbol{\Omega})=r$, $0<r<l$, then there is cointegration
and the matrix $\boldsymbol{\Omega}$ can be expressed as
$\boldsymbol{\Omega}=\boldsymbol{\alpha\beta}^\top$, where $\boldsymbol{\alpha}$
and $\boldsymbol{\beta}$ are
$l\times r$ matrices and
$\text{rank}(\boldsymbol{\alpha})=\text{rank}(\boldsymbol{\beta})=r$.
\item
The columns of $\boldsymbol{\beta}$ contains the cointegration vectors and the rows of
$\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
$\boldsymbol{\beta}$ is obtained by Johansen procedure~\cite{johansen1988},
whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.
\end{itemize}
It is worth noticing that the factorization of the matrix
$\boldsymbol\Omega$ is not unique, since for any $r \times r$
non-singular matrix $\mathbf{H}$, $\boldsymbol{\alpha}^*:=\boldsymbol{\alpha}\mathbf{H}$,
and $\boldsymbol{\beta}^*=\boldsymbol{\beta}(\mathbf{H}^{-1})^\top$ we have
$\boldsymbol{\alpha\beta}^\top=\boldsymbol{\alpha}^*(\boldsymbol{\beta}^*)^\top$.
If cointegration exists, then equation (\ref{eq:vec}) can be written
as follows:
\begin{equation}\label{eq:vecfull}
\Delta\mathbf{y}_t 
= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
which is a VAR model but for time series differences.

@@

\subsection{Ordinary Least Squares method}

The solution $\widehat{\mathbf{A}}$ to
equation~(\ref{eq:vareq}) can be obtained by the ordinary least squares (OLS)
method. $\widehat{\mathbf{X}}$ is the solution of the quadratic optimization problem
\begin{equation*}
\widehat{\mathbf{X}} = \underset{\mathbf{X}}{\text{Arg\;min}}
\|\mathbf{A}\mathbf{X}-\mathbf{B}\|_2^2
\end{equation*}
\noindent for which the solution $\widehat{\mathbf{X}}$ is well-known:
\begin{equation*}
\label{eq:MP}
\widehat{\mathbf{X}}=\mathbf{A}^{\!\!+}\,\mathbf{B}
\end{equation*}
\noindent where $\mathbf{A}^{\!\!+}$ is the Moore-Penrose pseudo-inverse
which, when $\mathbf{A}$ is full rank, can be written as follows: 
\begin{equation}
\label{eq:pseudoinverse}
\mathbf{A}^{\!\!+}= (\mathbf{A}^{\!\!\top} \mathbf{A})^{-1}\mathbf{A}^{\!\!\top} \, .
\end{equation}
when $\mathbf{A}$ is not full rank, i.e
$rank(\mathbf{A})=k <  n \leq m$, $\mathbf{A}^\top \mathbf{A}$ is
always singular and equation~(\ref{eq:pseudoinverse}) cannot be used.
More generally, the pseudo-inverse is best computed using the compact
singular value decomposition (SVD) of $\mathbf{A}$ which is:
\begin{equation*}
    \label{eq:compactsvd}
    \mathbf{A}=
    \mathbf{U_1}
    \boldsymbol \Sigma_1
    \mathbf{V}_1^{\top} \, .
\end{equation*}
Pseudo-inverse can then be written as follows:
\begin{equation*}
\label{eq:pseudoinversesvd}
\mathbf{A}^{\!\!+} = \mathbf{V}_1 \boldsymbol \Sigma_1^{-1} \mathbf{U}_1^\top \, .
\end{equation*}

\section{Methodology} \label{sec:methodology}
\subsection{Motivation} \label{sec:proposal}

Cointegration vectors can be found applying the Johansen method which uses a
sample of the last historical data. However, one of the main limitations of this
methodology is that it assumes cointegration vectors do not change in time.
In fact, the long-run relationships between the time series might change due to
several economic factors that can lead to structural breaks in the cointegrating
relation \cite{gregoryETal1996}. 
Experiments show that cointegration depends on the amount $L$ of historical data
considered, and how many lags $p$ we use.

Figure~\ref{fig:hists} shows the number $r$ of cointegration vectors given by
the Johansen method for different values of the parameters $L$ and $p$, using
four currencies data, i.e., $l=4$, and considering 1000 iterations $it$.
From the previous section \ref{sec:background} we know that when $r=0$, 
there is no cointegration and $r=l$ reveals that no process is I(1) but,
instead, they are all stationary.
The interesting cases of cointegration are those, where $r$ lies strictly
between $0$ and $l$, i.e. $0<r<l$,

@@

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.45\textwidth]{histCointVectors}
  \caption{Histogram of the number of cointegration vectors using $p=1$. Four
  possible values for $L$ are shown (100, 1900, 6100 and 10700).
  In the cases of $L=100$ and $L=6100$, most of the time there is little
  cointegration (0 or 4 cointegration vectors). However, for $L=1900$ and
  $L=10700$ significant cointegration is found.}
  \label{fig:hists}
\end{figure}

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.45\textwidth]{histCointVectorsp5}
  \caption{Histogram of the number of cointegration vectors using $p=5$. 
   Despite the chande in the distribution of cointegration vectors, 
   the percentage of cointegration is nearly the same as in the case $p=1$. 
   For example, for $L=10700$ the sum of 1 and 2 number of cointegration
   vectors is very similar to 1 cointegration vector for the case $L=10700$
   in figure \ref{fig:hists}.}
   \label{fig:histsp5}
\end{figure}

The same experiments, using now $p=5$, were carried out in order to see the
effect on the number of cointegration vectors found. 
Figure~\ref{fig:histsp5} shows that, despite the change in the distribution
of the number of cointegration vectors, the sum of their number with $0<r<4$,
remain nearly the same. 
Therefore we can say that the number of lags doesn't significantly affect
the extent of cointegration found.

@@

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.48\textwidth]{MAPEvsCoint-offset21600-p-2-freq-10s}
  \caption{MAPE versus percentage of cointegration. MAPE was obtained from 1000
  iterations considering several possible $L=[100,10000,400]$ and $p=[1,5]$.
  Best $L$ found was 8900.}
  \label{fig:cointvsmape}
\end{figure}

From these results we observe that cointegration changes with different values
for $L$ and $p$ and also that better cointegration percentage leads to better
accuracy performance.
Therefore we propose to choose best $L$ and $p$ parameters determined at every
step. 
The criterion for choosing $L$ and $p$ is to maximize the percentage of
cointegration $PC$ considering a number of past iterations $it$.

Algorithm ~\ref{alg:AVECM} summarizes our proposal. 
The function {\bf get\_best\_params} makes a grid search on the two vector
lists $Ls$ and $ps$ and returns the parameters $L$ and $p$ which maximize
the percentage of cointegration $PC$ (see equation~\ref{eq:pcoint}) for a
pre-defined number of iterations.
Since this search is computational expensive, the number of cointegration
vectors at each iteration is computed in parallel, thus ensuring a response
before new data is available. 
This parallel routine also allows to refine our grid search leading to obtain
improved forecasting performance.
After that optimum $L$ and $p$ parameters are found, VECM is built and used
to forecast next data.

@@

\subsection{Evaluation methods} \label{sec:evaluation}

Forecast performance was evaluated using different methods. We chose three
measures that are frequently used:
\begin{description}
\item
{\bf MAPE}, Mean Average Percent Error, which presents forecast errors as a
percentage:
\begin{equation}\label{eq:MAPE}
\text{MAPE} = \frac{1}{N} \sum_{t=1}^{N} 
\frac{\left|\mathbf{y}_t-\hat{\mathbf{y}}_t\right|}{\left|\mathbf{y}_t\right|}
 \times 100 
\end{equation}
\item
{\bf MAE}, Mean Average Error, which measures the distance between forecasts and the
true value.
\begin{equation}\label{eq:MAE}
\text{MAE} = \frac{1}{N} \sum_{t=1}^{N} 
\left| 
\mathbf{y}_t-\hat{\mathbf{y}}_t
\right| 
\end{equation}
\item
{\bf RMSE}, Root Mean Square Error, also measures the distance between forecasts
and the true values but, unlike MAE, large deviations from the true value have a
large impact on RMSE due to squaring forecast error.
\begin{equation}\label{eq:RMSE}
\text{RMSE} = \sqrt{
\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}}
\end{equation}
\end{description}

@@

\begin{table}[h!]
\label{tab:adf}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
& \textbf{Statistic} & \textbf{Critical value} & \textbf{Result}\\
\hline
EURUSD          &  -0.05398   & -1.94101 & True       \\
$\Delta$ EURUSD & -89.12344   & -1.94101 & False       \\
GBPUSD          &  -0.88208   & -1.94101 & True          \\
$\Delta$ GBPUSD & -82.56501   & -1.94101 & False       \\
CHFUSD          & -0.49941    & -1.94101 & True         \\
$\Delta$ CHFUSD & -74.21940   & -1.94101 & False       \\
JPYUSD          &  0.36721    & -1.94101 & True        \\
$\Delta$ JPYUSD & -101.15589  & -1.94101 & False     \\ 
\hline
\end{tabular}
\end{center}
\caption{Unit roots tests for EURUSD, GBPUSD, USDCHF and USDJPY at 10-seconds
frequency.}
\end{table}

\subsection{Parallel implementation} \label{sec:paralell}

@@

\begin{figure}[!h]
  \centering
  \includegraphics[width=0.48\textwidth]{accuracy}
  \caption{AVECM out-of-sample forecasting for four currency rates.}
  \label{fig:accuracy}
\end{figure}

\begin{table*}[ht!]
\caption{Performance and execution times for min MAPE criteria}
\label{tab:stats}
\begin{center}
\begin{tabular}{|l|l|l|c|c|c|c|c|}
\hline
\multicolumn{3}{|c}{Parameters} & 
\multicolumn{3}{|c}{Out-of-sample Performance} &
\multicolumn{2}{|c|}{Time[s]} \\ 
\hline

@@

\begin{figure}[!h]
  \centering
  \includegraphics[width=0.5\textwidth]{extimes}
  \caption{Computing time of sequential and parallel algorithm is shown in the
  upper figure. Speed-up is shown below.}
  \label{fig:extimes}
\end{figure}

Execution times do not consider the loading data time, just the finding of
best parameters and the building the VECM model. The time MPI spends
transferring data and synchronizing process is about two seconds independently
of the number of process considered.
